A Formal Verification Framework for Smart Contracts Based on Symbolic Execution

### 2024-01-08 papers/static-checker/ecoop2024/ecoop2024.tex
revised symbolic execution

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain
addresses). The type system is simple in the sense that it does not
support quantification over types.

@@

\section {A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}

@@

\section{Symbolic Execution Model}
\label{sec:symbolic-execution-model}
\subsection{System Model}
Michelson serves as the low-level smart contract language for the Tezos blockchain platform. The processing and manipulation of data during the execution of smart contracts are managed using a stack. The representation of a stack can be described as follows:
\begin{itemize}
\item[]  \ensuremath{[\ ]} represents the empty stack.
\item[]  \ensuremath{hd :: tl} denotes a stack with the top element as \ensuremath{hd} and the remaining stack as \ensuremath{tl}.
\end{itemize}
\begin{figure} []
\begin{align*}
T, U &::= \\
   &\Mid\ \langle \text{comparable type}\rangle \\
   &\Mid\ \text{option}\ \langle\text{type}\rangle \\
   &\Mid\ \text{list}\ \langle\text{type}\rangle \\
   &\Mid\ \text{set}\ \langle\text{comparable type}\rangle \\
   &\Mid\ \text{operation} \\
   &\Mid\ \text{contract}\ \langle\text{type}\rangle \\
   &\Mid\ \text{ticket}\ \langle\text{comparable type}\rangle \\
   &\Mid\ \text{pair}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{or}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{lambda}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{map}\ \langle\text{comparable type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{big-map}\ \langle\text{comparable type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{bls12-381-g1} \\
   &\Mid\ \text{bls12-381-g2} \\
   &\Mid\ \text{bls12-381-fr} \\
   &\Mid\ \text{sapling-transaction}\ \langle\text{natural number constant}\rangle \\
   &\Mid\ \text{sapling-state}\ \langle\text{natural number constant}\rangle \\
   &\Mid\ \text{chest} \\
   &\Mid\ \text{chest-key} \\
\langle\text{comparable type}\rangle ::= \\
   &\Mid\ \text{unit} \\
   &\Mid\ \text{never} \\
   &\Mid\ \text{bool} \\
   &\Mid\ \text{int}\\
   &\Mid\ \text{nat}\\
   &\Mid\ \text{string}\\
   &\Mid\ \text{chain-id}\\
   &\Mid\ \text{bytes}\\
   &\Mid\ \text{mutez}\\
   &\Mid\ \text{key-hash}\\
   &\Mid\ \text{key}\\
   &\Mid\ \text{signature}\\
   &\Mid\ \text{timestamp}\\
   &\Mid\ \text{address}\\
   &\Mid\ \text{tx-rollup-l2-address}\\
   &\Mid\ \text{option}\ \langle\text{comparable type}\rangle\\
   &\Mid\ \text{or}\ \langle\text{comparable type}\rangle \langle\text{comparable type}\rangle\\
   &\Mid\ \text{pair}\ \langle\text{comparable type}\rangle \langle\text{comparable type}\rangle \DOT \\
\end{align*}
\caption{Types}
\label{fig:type}
\end{figure}
Each piece of data in Michelson is assigned a specific type, and the language restricts programmers from introducing custom types. The complete list of types can be found in Figure \ref{fig:type}. To model the system, we use a stack in which each element is a pair consisting of term and its corresponding type. Terms are defined in Figure \ref{fig:term}. The stack is denoted as \STACK\ = (\TermOne, \TYF) \STACKCONCAT\ (\TermTwo, \TYS) \STACKCONCAT\ \DOT\ \STACKCONCAT\ \EMPTYSTACK\ where elements are types paired with terms.

A Michelson program is a sequence of instructions executed in a sequential manner. Each instruction takes, as input, the stack produced by the preceding instruction and modifies it for the subsequent one. Let \INSTRUCTION\ = \InstructionOne; \InstructionTwo; \DOT\  \InstructionN\ represent a sequence of instructions.
During symbolic execution,  the system state evolves, assuming different values based on the constraints of symbolic values. Moreover, branching may occur depending on conditions, and both constraints and branching conditions are recorded as predicates. Let \PREDICATE\ is a path predicate that is extended at each conditional and expressed in conjunction form, capturing the branching conditions and constraints encountered during the symbolic execution process.
\begin{definition}
A symbolic execution state is represented by the tuple \STATE\ = [\INSTRUCTION, \STACK, \PREDICATE]. Here, \SYSTEM\ = \{\STATEONE, \STATETWO, \DOT, \STATEN \} range over sets of  states.
\end{definition}

The smart contract program takes a stack as input, consisting of a single pair where the first element is an input value referred to as parameter denoted as \VPAR and the second element represents the content of the storage space denoted as \VSTORAGE. The program produces a stack as output, which is a single pair. The first element of this pair is the list of internal operations that the program intends to emit (\VOPERATIONLIST), and the second element is the new contents of the storage space.  Let's define the initial stack as  \SINIT\   = (\KPAIR\ \VPAR\ \VSTORAGE, \TPAIR\ \TYF\ \TYS) \STACKCONCAT\ \EMPTYSTACK, where \VPAR\ and \VSTORAGE\ denote terms representing the parameter and the storage, and \TYF\ and \TYS\ are their corresponding types. Similarly, designate the final stack as \SFINAL\   = (\PAIR\ \VOPERATIONLIST\ \VSTORAGE, \TPAIR\ (\TOPERATIONLIST) \TYS) \STACKCONCAT\ \EMPTYSTACK, where \VOPERATIONLIST\ repesents a operation list.
\begin{figure}[]
\begin{align*}
\text{t} &::= \\
   &\Mid\ \langle \text{variable} \rangle \\
   &\Mid\ \langle \text{account constant} \rangle \\
   &\Mid\ \langle \text{int constant} \rangle \\
   &\Mid\ \langle \text{string constant} \rangle \\
   &\Mid\ \langle \text{byte sequence constant} \rangle \\
   &\Mid\ \UNIT \\
   &\Mid\ \NEVER \\
   &\Mid\ \TRUE \\
   &\Mid\ \FALSE \\
   &\Mid\ \PAIR\ \text{t1 t2} \\
   &\Mid\ \LEFT\ \text{t}\\
   &\Mid\ \RIGHT\ \text{t}\\
   &\Mid\ \SOME\ \text{t}\\
   &\Mid\ \NONE \\
   &\Mid\ \text{\{t ; ... \}}\\
   &\Mid\ \text{\{ Elt t1 t2 ; ... \}}\\
   &\Mid\ \{ \langle\text{instruction}\rangle; ... \}   \\
\langle \text{variable} \rangle &::= \\ 
   &\Mid\ \VAR\\
\langle \text{account constant} \rangle &::= \\ 
   &\Mid\ \text{balance} \\
   &\Mid\ \text{amount} \\
   &\Mid\ \text{sender} \\
   &\Mid\ \text{source} \\
   &\Mid\ \text{now} \\
   &\Mid\ \text{level} \\
   &\Mid\ \text{chain-id} \\
   &\Mid\ \text{self}  \\
   &\Mid\ \text{self-address}  \\
   &\Mid\ \text{total-voting-power}  \\
   &\Mid\ \text{voting-power}  \\
\langle\text{instruction}\rangle &::= \\
  &\Mid\ \DROP \\
  &\Mid\ \DROP \langle\text{natural number constant}\rangle \\
  &\text{...}
\end{align*}
\caption{Terms}
\label{fig:term}
\end{figure}
\subsection{Rules}
Each instruction is defined by a rule on the system model. The rule semantic is defined by several kinds of transitions:
\begin{enumerate}
\item \ExprTrans\ single-step evaluation of an expression in a system state,
\item \StateTrans\ internal transitions of a state, which is non-deterministic  as shown for loop instructions, such as \LOOP\ and \LOOPLEFT,
\item \SystemTrans\ symbolic system transitions.
\end{enumerate}

During the symbolic execution, numerous states are generated, but some may be unreachable from a symbolic initial state. This happens when the predicate \PREDICATE\ is unsatisfied, leading to the exclusion of a system state where the predicate fails.

\begin{mathpar}
\inferrule[]
  { \NEG\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}
Michelson instructions are categorized based on their functions. These include such as control structures, stack manipulation, arithmetic operations, boolean operations, cryptographic operations, and operations on data structures. In our model, we classified them as one-step, multi-step, blockchain, cryptographic, branch, and loop instructions. This categorization reflects how these instructions are handled within our specific model.
\subsubsection{One-step Instructions}
One-step instructions have a localized effect, modifying the system state directly without branching a new state, and they end the execution after a single operation. The ABS rule exemplifies such an instruction, operating on the first element of the stack. This element must be of type int, and the instruction returns the positive value of its current value, treated as a natural number.
%ABS
\begin{mathpar}
\inferrule[ABS]
  {
  }
  {
    [(\ABS; \INSTRUCTION), (\StackOne, \TINT) \STACKCONCAT \STACK,
    \PREDICATE]
    \StateTrans \\
    [\INSTRUCTION, (\X, \TNAT) \STACKCONCAT \STACK,
    \PREDICATE\ \wedge\ (\StackOne\ \GE\ \ZERO\ \Rightarrow\ \X\ =
    \StackOne) \wedge\ (\StackOne\ \LT\ \ZERO\ \Rightarrow\ \X\ = \MINUS\ \StackOne)]
 }
\end{mathpar}
Another example is the \COMPARE\ instruction for natural numbers, which compares the first two elements of the stack, both having a natural number type. It returns a result of either -1, 0, or 1 based on the comparison of their values.
% COMPARE
\begin{mathpar}
\inferrule[COMPARE]
  {
  }
  {
    [(\COMPARE ; \INSTRUCTION), (\StackOne, \TNAT) \STACKCONCAT (\StackTwo, \TNAT)
    \STACKCONCAT \STACK, \PREDICATE ]
    \SystemTrans \\
    [\INSTRUCTION, (\X, \TINT) \STACKCONCAT \STACK, \PREDICATE
    \wedge\ (\StackOne\ \GT\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \ONE)
    \wedge\ (\StackOne\ \EQ\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \ZERO) 
    \wedge\ (\StackOne\ \LT\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \MINUS \ONE)]
    }
\end{mathpar}
\subsubsection{Multi-step Instructions}
Multi-step instructions involve the execution of a sub-sequence of instructions before returning to the main execution. The rule below models the \EXEC\ instruction, which executes a function specified as a sequence of instructions denoted by \INSTRUCTIONONE\ in the rule. This instruction applies the code of the function to the first element of the stack and then places the result back into the main execution.
\begin{mathpar}
  \inferrule[\EXEC]
  {
    [\INSTRUCTIONONE, (\StackOne, \TYF) \STACKCONCAT \EMPTYSTACK, 
    \Q]
    \StateTrans^*
    [ \EMPTYSTACK, (\StackOne', \TYS) \STACKCONCAT \Q']
  }{
     [(\EXEC; \INSTRUCTION),   (\{\INSTRUCTIONONE\}, \TYF\ \rightarrow\ \TYS) \STACKCONCAT
    (\StackOne, \TYF) \STACKCONCAT \STACK, \PREDICATE \wedge \Q] 
    \StateTrans \
    [ \INSTRUCTION, (\StackOne', \TYS) \STACKCONCAT \STACK,
    \PREDICATE \wedge \Q']}
\end{mathpar}

The \DIP\ \N\ instruction delves into the first \N\ elements of a stack, applies the sequence of instructions denoted as \INSTRUCTIONONE\ to the remaining part of the stack, modifies a part of the predicate and subsequently pushes the result back into the main execution.
%DIP n
\begin{mathpar}
\inferrule[\DIP\ \N]
  { 
     \FLEN(\A) \EQ\ \N \\ [\INSTRUCTIONONE,  \B, \Q]
    \StateTrans^*
    [\EMPTYSTACK,  \B_1, \Q']
  }
  {[(\DIP\ \N\ \INSTRUCTIONONE; \INSTRUCTION), \A\ \At\ \B, \PREDICATE \wedge \Q] \StateTrans 
[(\INSTRUCTION), \A\ \At\ \B_1, \PREDICATE \wedge \Q']}
\end{mathpar}
\subsubsection{Blockchain Instructions}
 Blockchain instructions make implicit use of data from an execution environment, derived from the current state of the blockchain (the context), from the transaction that triggered the contract, and from the block containing this transaction. We model these data as symbolic values, which satisfis some constrants. For example, the \AMOUNT\ instruction adds to the stack the amount of tokens that sent together with the transcation which calls this smart contract. Here, the amount is modeled as a symbolic value donated as \CAMOUNT, which has \TMUTEZ\ type, which has a nature number value.

@@

\subsubsection{Cryptographic Instructions}
Cryptographic instructions in Michelson utilize built-in functions that manage cryptographic features. These functions are symbolically modeled in our framework, specifying their inputs and outputs. For instance, the instruction \SHA\ computes the cryptographic hash of the top of the stack using the SHA-256 cryptographic hash function. The function SHA-256 is symbolically represented as the function  \FSHA\  that takes an input of type \TBYTE\ and returns another byte value, constrained to a length of 256.
%HASH-KEY
\begin{mathpar}
\inferrule[\SHA]
  {
  }
  {[(\SHA; \INSTRUCTION), (\StackOne, \TBYTE) \STACKCONCAT\STACK, \PREDICATE] \StateTrans \
[\INSTRUCTION, (\X, \TBYTE) \STACKCONCAT\STACK, \PREDICATE \Wedge\ (\X\ \EQ\ \FSHA(\StackOne))]}
\end{mathpar}
\subsubsection{Branch Instructions}
The Michelson language incorporates conditional branching, enabling the execution of specific code segments based on certain criteria. One of the instructions facilitating this is the \IF\ \{\} \{\} construct. For example, the \IF\ \{ \INSTRUCTIONONE\ \} \{ \INSTRUCTIONTWO\  \} instruction allows the creation of branches in execution. It takes two sequences as arguments. Expecting a boolean value at the top of the stack, it consumes this top element. If the boolean is True, it executes the first provided sequence; otherwise, it executes the second sequence.
%IF
\begin{mathpar}
  \inferrule[IF-true]
  {  
  }{
    [(\IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TBOOL) \STACKCONCAT\STACK, \PREDICATE]
    \StateTrans\
    [\INSTRUCTIONONE, \STACK, \PREDICATE\ \Wedge\ \StackOne]
  }

@@

Another branch instruction is the IF-LEFT instruction, which is used for conditional execution based on the left or right branch of a tagged union (sum) type. 
%IF-LEFT-LEFT
\begin{mathpar}
  \inferrule[IF-LEFT-left]
  {  
  }{
    [(\IFLEFT\ \INSTRUCTIONONE\ \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TOR\ \TYF\ \TYS) \STACKCONCAT \STACK, \PREDICATE]
    \StateTrans \
    [\INSTRUCTIONONE, (\X, \TYF) \STACKCONCAT\STACK,
    \PREDICATE \wedge (\StackOne\ \EQ\ \LEFT\ \X)]
  }
\end{mathpar}

%IF-LEFT-RIGHT
\begin{mathpar}
  \inferrule[IF-LEFT-right]
  {  
  }{
    [(\IFLEFT\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TOR\ \TYF\ \TYS) \STACKCONCAT \STACK, \PREDICATE]
    \StateTrans \
    [\INSTRUCTIONTWO, (\X, \TYS) \STACKCONCAT\STACK, \PREDICATE \wedge (\StackOne\ \EQ\ \RIGHT\ \X))]
  }
\end{mathpar}
These above rules collectively define the behavior of the \IFLEFT instruction based on the left or right branch of a tagged union type. This rule applies when the top element of the stack is a tagged union of type \TOR\ \TYF\ \TYS (either of type \TYF\ or \TYS), and the \IFLEFT instruction is followed by instructions \INSTRUCTIONONE\ and \INSTRUCTIONTWO. The first rule apply with the condition that the top element of the stack is of the form Left \X\ (indicating the left branch of the union), it transitions to execute the instructions \INSTRUCTIONONE\ with the stack modified accordingly and the predicate is updated to include the condition that the top element of the stack is equal to Left \X.  Similerly, the second rule apply with the condition that  the top element is of the form Right \X\ (indicating the right branch of the union), it transitions to execute the instructions \INSTRUCTIONTWO\ with the stack modified accordingly and the predicate is updated to include the condition that the top element of the stack is equal to Right \X.

>>>>

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions)
with the expected operations,
Michelson supports a number of types and operations specific to blockchain applications (e.g.,
\texttt{mutez} for tokens, \texttt{address} for blockchain
addresses). The type system is simple in the sense that it does not
support quantification over types. There are no user-defined types.

@@

\section{A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}

@@

\section{Symbolic Execution Model}
\label{sec:symbolic-execution-model}

We already learned about Michelson, the low-level smart contract
language for the Tezos blockchain, and its stack-based nature in
Section~\ref{sec:background}.  Here we consider the changes needed for
symbolic execution of Michelson programs.

\subsection{System Model}
\label{sec:system-model}
\begin{figure}[tp]
\begin{align*}
T, U &::= \\
   &\Mid\ \langle \text{comparable type}\rangle \\
   &\Mid\ \text{option}\ \langle\text{type}\rangle \\
   &\Mid\ \text{list}\ \langle\text{type}\rangle \\
   &\Mid\ \text{set}\ \langle\text{comparable type}\rangle \\
   &\Mid\ \text{operation} \\
   &\Mid\ \text{contract}\ \langle\text{type}\rangle \\
   &\Mid\ \text{ticket}\ \langle\text{comparable type}\rangle \\
   &\Mid\ \text{pair}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{or}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{lambda}\ \langle\text{type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{map}\ \langle\text{comparable type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{big-map}\ \langle\text{comparable type}\rangle \langle\text{type}\rangle \\
   &\Mid\ \text{bls12-381-g1} \\
   &\Mid\ \text{bls12-381-g2} \\
   &\Mid\ \text{bls12-381-fr} \\
   &\Mid\ \text{sapling-transaction}\ \langle\text{natural number constant}\rangle \\
   &\Mid\ \text{sapling-state}\ \langle\text{natural number constant}\rangle \\
   &\Mid\ \text{chest} \\
   &\Mid\ \text{chest-key} \\
\langle\text{comparable type}\rangle ::= \\
   &\Mid\ \text{unit} \\
   &\Mid\ \text{never} \\
   &\Mid\ \text{bool} \\
   &\Mid\ \text{int}\\
   &\Mid\ \text{nat}\\
   &\Mid\ \text{string}\\
   &\Mid\ \text{chain-id}\\
   &\Mid\ \text{bytes}\\
   &\Mid\ \text{mutez}\\
   &\Mid\ \text{key-hash}\\
   &\Mid\ \text{key}\\
   &\Mid\ \text{signature}\\
   &\Mid\ \text{timestamp}\\
   &\Mid\ \text{address}\\
   &\Mid\ \text{tx-rollup-l2-address}\\
   &\Mid\ \text{option}\ \langle\text{comparable type}\rangle\\
   &\Mid\ \text{or}\ \langle\text{comparable type}\rangle \langle\text{comparable type}\rangle\\
   &\Mid\ \text{pair}\ \langle\text{comparable type}\rangle \langle\text{comparable type}\rangle \DOT \\
\end{align*}
\caption{Types}
\label{fig:type}
\end{figure}

Our symbolic model of a program state is a stack in which each element
is a symbolic value, i.e., a pair consisting
of term and its corresponding type. Terms are defined in
Figure~\ref{fig:term}. Thus we write a stack as \STACK\ = (\TermOne,
\TYF) \STACKCONCAT\ (\TermTwo, \TYS) \STACKCONCAT\ \DOT\ \STACKCONCAT\ 
\EMPTYSTACK\ where \ensuremath{[\ ]} stands for the empty stack and
\ensuremath{hd :: tl} denotes a stack with the top element as
\ensuremath{hd} and the remaining stack as \ensuremath{tl}. 
A selection of types is shown in Figure~\ref{fig:type}.

A Michelson program is a sequence of instructions executed in a
sequential manner. Each instruction takes, as input, the stack
produced by the preceding instruction and modifies it for the
subsequent one. Let \INSTRUCTION\ = \InstructionOne; \InstructionTwo;
\DOT\  \InstructionN\ represent a sequence of instructions.

During symbolic execution,  the system state evolves considering all
values admitted by the constraints on symbolic values. Unless the
input values are sufficiently known, symbolic execution may split the
state, and both constraints and branching conditions are recorded as
predicates. Let \PREDICATE\ be a path predicate that is extended at
each conditional and expressed in conjunctive form, capturing the
branching conditions and constraints encountered during the symbolic
execution process. 
\begin{definition}
A symbolic execution state is represented by a tuple \STATE\ =
[\INSTRUCTION, \STACK, \PREDICATE].
A symbolic system configuration, \SYSTEM\ = \{\STATEONE, \STATETWO,
\DOT, \STATEN \}, is a set of  states.
\end{definition}

The smart contract program takes a stack as input, consisting of a
single pair where the first element is an input value referred to as
parameter denoted as \VPAR\ and the second element represents the
content of the storage space denoted as \VSTORAGE. The program
produces a stack as output, which is a single pair. The first element
of this pair is the list of internal operations that the program
intends to emit (\VOPERATIONLIST), and the second element is the new
contents of the storage space.  We define the initial stack as
\SINIT\   = (\KPAIR\ \VPAR\ \VSTORAGE, \TPAIR\ \TYF\ \TYS)
\STACKCONCAT\ \EMPTYSTACK, where \VPAR\ and \VSTORAGE\ denote terms
representing the parameter and the storage, and \TYF\ and \TYS\ are
their corresponding types. Similarly, a final stack has the form
\SFINAL\   = (\PAIR\ \VOPERATIONLIST\ \VSTORAGE, \TPAIR\
(\TOPERATIONLIST) \TYS) \STACKCONCAT\ \EMPTYSTACK, where
\VOPERATIONLIST\ repesents a operation list.

The initial configuration is \SYSTEMINIT\ = \{[\INSTRUCTION, \SINIT, \TRUE] \}. 
\begin{figure}[]
\begin{align*}
\text{t} &::= \\
   &\Mid\ \langle \text{variable} \rangle \\
   &\Mid\ \langle \text{account constant} \rangle \\
   &\Mid\ \langle \text{int constant} \rangle \\
   &\Mid\ \langle \text{string constant} \rangle \\
   &\Mid\ \langle \text{byte sequence constant} \rangle \\
   &\Mid\ \UNIT \\
   &\Mid\ \NEVER \\
   &\Mid\ \TRUE \\
   &\Mid\ \FALSE \\
   &\Mid\ \PAIR\ \text{t1 t2} \\
   &\Mid\ \LEFT\ \text{t}\\
   &\Mid\ \RIGHT\ \text{t}\\
   &\Mid\ \SOME\ \text{t}\\
   &\Mid\ \NONE \\
   &\Mid\ \text{\{t ; ... \}}\\
   &\Mid\ \text{\{ Elt t1 t2 ; ... \}}\\
   &\Mid\ \{ \langle\text{instruction}\rangle; ... \}   \\
\langle \text{variable} \rangle &::= \\ 
   &\Mid\ \VAR\\
\langle \text{account constant} \rangle &::= \\ 
   &\Mid\ \text{balance} \\
   &\Mid\ \text{amount} \\
   &\Mid\ \text{sender} \\
   &\Mid\ \text{source} \\
   &\Mid\ \text{now} \\
   &\Mid\ \text{level} \\
   &\Mid\ \text{chain-id} \\
   &\Mid\ \text{self}  \\
   &\Mid\ \text{self-address}  \\
   &\Mid\ \text{total-voting-power}  \\
   &\Mid\ \text{voting-power}  \\
\langle\text{instruction}\rangle &::= \\
  &\Mid\ \DROP \\
  &\Mid\ \DROP \langle\text{natural number constant}\rangle \\
  &\text{...}
\end{align*}
\caption{Terms}
\label{fig:term}
\end{figure}
\subsection{Rules}
Each instruction is defined by a rule on a system configuration. There
are several kinds of transitions:
\begin{enumerate}
% \item \ExprTrans\ single-step evaluation of an expression in a system state,
\item \StateTrans\ internal transitions of a state, which is
  non-deterministic  as shown for loop instructions, such as \LOOP\
  and \LOOPLEFT, 
\item \SystemTrans\ symbolic system transitions.
\end{enumerate}

A system transition picks any state and replaces it by all distinct states that
are reachable by internal transitions. It exploits the nondeterminism
of the \StateTrans\ relation.
\begin{mathpar}
  \inferrule[]
  { \STATE\ \StateTrans \STATEONE, \dots, \STATE\ \StateTrans \STATEN
  }{
  \{\STATE\} \cup \SYSTEM \SystemTrans \{ \STATEONE, \dots, \STATEN \}
  \cup \SYSTEM}  
\end{mathpar}
In general, symbolic execution may generate unreachable
states. A state is unreachable if its predicate \PREDICATE\ is  not
satisfiable. Hence, we have a rule that drops unreachable states.
\begin{mathpar}
\inferrule[]
  { \UNSAT\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}
The working of the symbolic interpreter may also be understood as a
graph with symbolic states as nodes and internal transitions as
edges. Such a representation makes it easier to perform multi-step
subtransitions as they are required for some operations.

Michelson instructions are categorized based on their functions. The
categories include control structures, stack manipulation, arithmetic
operations, boolean operations, cryptographic operations, and
operations on data structures. In our model, we classify them as
one-step, multi-step, blockchain, cryptographic, branch, and loop
instructions. This categorization reflects how these instructions are
handled within our specific model.

\subsubsection{One-step Instructions}
\label{sec:one-step-instr}
One-step instructions have a localized effect, modifying the system
state directly without branching a new state, and their execution is
modeled with a single rule.
% The ABS rule exemplifies such an
% instruction, operating on the first element of the stack. This element
% must be of type int, and the instruction returns the absolute value of
% its argument as a natural number. To this end, the rule creates a
% fresh variable, \X, and defines its value by a suitable predicate.
% %ABS
% \begin{mathpar}
% \inferrule[ABS]
%   { \X\ \FRESH
%   }
%   {
%     [(\ABS; \INSTRUCTION), (\StackOne, \TINT) \STACKCONCAT \STACK,
%     \PREDICATE]
%     \StateTrans \\
%     [\INSTRUCTION, (\X, \TNAT) \STACKCONCAT \STACK,
%     \PREDICATE\ \wedge\ (\StackOne\ \GE\ \ZERO\ \Rightarrow\ \X\ =
%     \StackOne) \wedge\ (\StackOne\ \LT\ \ZERO\ \Rightarrow\ \X\ = \MINUS\ \StackOne)]
%  }
% \end{mathpar}

For example, the \COMPARE\ instruction for natural numbers compares
the first two elements of the stack, both having a natural number
type. It returns a result of either -1, 0, or 1 based on the
comparison of their values. 
% COMPARE
\begin{mathpar}
\inferrule[COMPARE]
  { \X\ \FRESH
  }
  {
    [(\COMPARE ; \INSTRUCTION), (\StackOne, \TNAT) \STACKCONCAT (\StackTwo, \TNAT)
    \STACKCONCAT \STACK, \PREDICATE ]
    \SystemTrans \\
    [\INSTRUCTION, (\X, \TINT) \STACKCONCAT \STACK, \PREDICATE
    \wedge\ (\StackOne\ \GT\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \ONE)
    \wedge\ (\StackOne\ \EQ\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \ZERO) 
    \wedge\ (\StackOne\ \LT\ \StackTwo\ \Leftrightarrow\ \X\ \EQ\ \MINUS \ONE)]
    }
\end{mathpar}
\subsubsection{Multi-step Instructions}
Multi-step instructions involve the execution of a sub-sequence of instructions before returning to the main execution. The rule below models the \EXEC\ instruction, which executes a function specified as a sequence of instructions denoted by \INSTRUCTIONONE\ in the rule. This instruction applies the code of the function to the first element of the stack and then places the result back into the main execution.
\begin{mathpar}
  \inferrule[\EXEC]
  {
    [\INSTRUCTIONONE, (\StackOne, \TYF) \STACKCONCAT \EMPTYSTACK, 
    \Q]
    \StateTrans^*
    [ \EMPTYSTACK, (\StackOne', \TYS) \STACKCONCAT \Q']
  }{
     [(\EXEC; \INSTRUCTION),   (\{\INSTRUCTIONONE\}, \TYF\ \rightarrow\ \TYS) \STACKCONCAT
    (\StackOne, \TYF) \STACKCONCAT \STACK, \PREDICATE \wedge \Q] 
    \StateTrans \
    [ \INSTRUCTION, (\StackOne', \TYS) \STACKCONCAT \STACK,
    \PREDICATE \wedge \Q']}
\end{mathpar}

The \DIP\ \N\ instruction skips over the first \N\ elements of a
stack, applies the sequence of instructions denoted as
\INSTRUCTIONONE\ to the remaining part of the stack, modifies a part
of the predicate and subsequently pushes the result back into the main
execution. 
%DIP n
\begin{mathpar}
\inferrule[\DIP\ \N]
  { 
     \FLEN(\A) \EQ\ \N \\ [\INSTRUCTIONONE,  \B, \Q]
    \StateTrans^*
    [\EMPTYSTACK,  \B_1, \Q']
  }
  {[(\DIP\ \N\ \INSTRUCTIONONE; \INSTRUCTION), \A\ \At\ \B, \PREDICATE \wedge \Q] \StateTrans 
[\INSTRUCTION, \A\ \At\ \B_1, \PREDICATE \wedge \Q']}
\end{mathpar}
\subsubsection{Blockchain Instructions}
 Blockchain instructions make implicit use of data from the current
 state of the blockchain (the context), from the transaction that
 triggered the contract, and from the block containing this
 transaction. As these values are constant during the execution of a
 contract, we model them as fixed variable. For example, the \AMOUNT\
 instruction pushes the amount of tokens on the stack that the
 contract received from its caller. This amount is modeled as the
 fixed variable \CAMOUNT, which has type \TMUTEZ.

@@

% \subsubsection{Cryptographic Instructions}
% Cryptographic instructions in Michelson utilize built-in functions that manage cryptographic features. These functions are symbolically modeled in our framework, specifying their inputs and outputs. For instance, the instruction \SHA\ computes the cryptographic hash of the top of the stack using the SHA-256 cryptographic hash function. The function SHA-256 is symbolically represented as the function  \FSHA\  that takes an input of type \TBYTE\ and returns another byte value, constrained to a length of 256.
% %HASH-KEY
% \begin{mathpar}
% \inferrule[\SHA]
%   {
%   }
%   {[(\SHA; \INSTRUCTION), (\StackOne, \TBYTE) \STACKCONCAT\STACK, \PREDICATE] \StateTrans \
% [\INSTRUCTION, (\X, \TBYTE) \STACKCONCAT\STACK, \PREDICATE \Wedge\ (\X\ \EQ\ \FSHA(\StackOne))]}
% \end{mathpar}
\subsubsection{Branch Instructions}
The simplest conditional in Michelson language is the \IF\ \{
\INSTRUCTIONONE\ \} \{ \INSTRUCTIONTWO\  \} instruction. It expects
and consumes a boolean value at the top of the stack. If the boolean
is True, it executes \INSTRUCTIONONE; otherwise, it executes
\INSTRUCTIONTWO. 
%IF
\begin{mathpar}
  \inferrule[IF-true]
  {  
  }{
    [(\IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TBOOL) \STACKCONCAT\STACK, \PREDICATE]
    \StateTrans\
    [\INSTRUCTIONONE, \STACK, \PREDICATE\ \Wedge\ \StackOne]
  }

@@

Another branch instruction is the IF-LEFT instruction, which is used
for conditional execution based on the left or right branch of a sum
type.  
%IF-LEFT-LEFT
\begin{mathpar}
  \inferrule[IF-LEFT-left]
  {  \X\ \FRESH
  }{
    [(\IFLEFT\ \INSTRUCTIONONE\ \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TOR\ \TYF\ \TYS) \STACKCONCAT \STACK, \PREDICATE]
    \StateTrans \
    [(\INSTRUCTIONONE; \INSTRUCTION), (\X, \TYF) \STACKCONCAT\STACK,
    \PREDICATE \wedge (\StackOne\ \EQ\ \LEFT\ \X)]
  }
\end{mathpar}

%IF-LEFT-RIGHT
\begin{mathpar}
  \inferrule[IF-LEFT-right]
  {    \X\ \FRESH
  }{
    [(\IFLEFT\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TOR\ \TYF\ \TYS) \STACKCONCAT \STACK, \PREDICATE]
    \StateTrans \
    [(\INSTRUCTIONTWO; \INSTRUCTION), (\X, \TYS) \STACKCONCAT\STACK, \PREDICATE \wedge (\StackOne\ \EQ\ \RIGHT\ \X))]
  }
\end{mathpar}
These above rules define the behavior of the \IFLEFT instruction. The
rule applies when the top element of the stack is a tagged union of
type \TOR\ \TYF\ \TYS (either of type \TYF\ or \TYS), and the \IFLEFT
instruction is followed by instructions \INSTRUCTIONONE\ and
\INSTRUCTIONTWO. The first rule adds the condition that the top
element of the stack is of the form Left \X\ (indicating the left
branch of the union), it transitions to execute the instructions
\INSTRUCTIONONE\ with the stack modified accordingly and the predicate
is updated to include the condition that the top element of the stack
is equal to Left \X. Similarly, the second rule handles the case Right\
\X, mutatis mutandis.

\subsubsection{Loop Instructions}
Michelson supports a number of loop instructions, including
\LOOP\ and \LOOPLEFT, that  check the loop condition by examining the
top of the stack. These loop instructions may not terminate, just like while-loops. There are further loop instructions, like \ITER\ and
\MAP, specialized to traverse data structures like lists, sets, and maps. There are also
some instructions that implicitly require looping like \CONCAT, and
\SIZE. The latter kinds of loop instructions do always terminate.
